---
title: "sampling_analysis_SP_C_06_18_25"
format: html
---
###Load Libraries

```{r load libraries}
library(tidyverse)
library(ggplot2)
library(janitor)
library(stringr)
library(dplyr)
library(leaflet)
library(sf)
library(patchwork)
library(purrr)
library(lme4)
library(lmerTest)
library(dbscan)
```

### Load Data:

```{r data pull, echo=FALSE, message=FALSE, warning=FALSE}
# Load the data
lime <- read_csv("../data_do_not_push/2021 Liming Study Soil Data Compiled (lbs. A).csv") %>%
# Clean the data
  clean_names() %>%
  #split "sample" column by type and site (0-4,is type, Site 1,2... is site)
  separate(sample, into = c("type", "site"), sep = ", ") %>%
  #make sure site rows are lowercase and _ instead of spaces:
  mutate(site = tolower(site),
         site = str_replace_all(site, " ", "_")) %>%
  #group by site and the SAME lat long:
  group_by(site, lat, long) %>%
  arrange(lat, long) %>%
  #filter out 4-8 sample types:
  filter(type != "4-8") %>%
  mutate(location = cur_group_id()) %>%   # unique id for lat lon
  ungroup() %>%    
  arrange(location) %>%
  #add new column and copy over all 0-4 values, but change A and B to SP:
  mutate(sampling_type = case_when(
    type == "0-4" ~ "C",
    type == "A" ~ "SP",
    type == "B" ~ "SP",
    TRUE ~ type
  ))



```



#### Map Data:
```{r}

lime_sf <- lime %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) # WGS84 (lon/lat)

#create leaflet plot of lime_sf with points colored by type and labelled by location:

leaflet(lime_sf) %>%
  addTiles() %>%
  addCircleMarkers(
    color = ~case_when(
      type == "0-4" ~ "red",
      type == "A" ~ "green",
      type == "B" ~ "pink"),
    label = ~paste("Location:", location, "<br>Type:", type),
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.8
  ) %>%
  addLegend("bottomright", 
            colors = c("red", "green", "pink"),
            labels = c("0-4","A","B"),
            title = "Sample Type") %>%
  #change basemap to satellite:
  addProviderTiles("Esri.WorldImagery")
```
### Per site analysis

```{r Cluster by site}

# Step 1: Convert lime to sf (CRS 4326 for original lat/lon)
lime_sf <- lime %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326)

# Step 2: Transform to UTM (meters) for distance-based clustering
lime_utm <- st_transform(lime_sf, crs = 32617)  # UTM Zone 17N

# Step 3: Extract coordinates for DBSCAN
coords <- st_coordinates(lime_utm)

# Step 4: Run DBSCAN clustering — try smaller eps (in meters)
# eps = 10000 is 10km, adjust based on how spread out your points are
clustering <- dbscan(coords, eps = 10000, minPts = 2)

# Step 5: Attach clusters back to original (in degrees)
lime_sf$cluster <- as.factor(clustering$cluster)  # back in CRS 4326

# Step 6: Visualize in Leaflet
leaflet(lime_sf) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>%
  addCircleMarkers(
    radius = 5,
    color = ~colorFactor(rainbow(length(unique(lime_sf$cluster))), lime_sf$cluster)(cluster),
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste("Cluster:", cluster)
  )


```
```{r log transformed (aside from pH) and analyzed based on clustered sites}
# Define which nutrients to log-transform
nutrients <- c("ca", "k", "mg", "mn", "p", "zn", "p_h_2", "lbc_1", "lb_ceq")
nutrients_to_log <- setdiff(nutrients, "p_h_2")

# Apply log1p transformation to selected columns
lime_sf <- lime_sf %>%
  mutate(across(all_of(nutrients_to_log), ~ log1p(.)))

# Run your statistical tests as before
run_tests_per_cluster <- function(df, nutrient) {
  df <- df %>% filter(!is.na(.data[[nutrient]]))
  if (n_distinct(df$sampling_type) < 2) return(NULL)

  normality <- df %>%
    group_by(sampling_type) %>%
    summarise(p = shapiro.test(.data[[nutrient]])$p.value, .groups = "drop")

  use_t <- all(normality$p > 0.05)

  test_res <- if (use_t) {
    t.test(df[[nutrient]] ~ df$sampling_type)
  } else {
    wilcox.test(df[[nutrient]] ~ df$sampling_type)
  }

  tibble(
    cluster = unique(df$cluster),
    nutrient = nutrient,
    test = ifelse(use_t, "t-test", "wilcox"),
    p_value = test_res$p.value,
    statistic = test_res$statistic,
    median_C = median(df[[nutrient]][df$sampling_type == "C"], na.rm = TRUE),
    median_SP = median(df[[nutrient]][df$sampling_type == "SP"], na.rm = TRUE)
  )
}

results_by_cluster <- lime_sf %>%
  group_by(cluster) %>%
  group_map(~ map_dfr(nutrients, run_tests_per_cluster, df = .x), .keep = TRUE) %>%
  bind_rows() %>%
  arrange(nutrient)

print(results_by_cluster)

# Prepare long data for plotting
lime_long <- lime_sf %>%
  st_drop_geometry() %>%
  pivot_longer(cols = all_of(nutrients),
               names_to = "nutrient",
               values_to = "value") %>%
  mutate(sampling_type = factor(sampling_type, levels = c("C", "SP")))

# Name the clusters
cluster_names <- c(
  "1" = "Lower Coastal Plains",
  "2" = "Upper Coastal Plains",
  "3" = "Piedmont",
  "4" = "Valley and Ridge"
)

lime_long <- lime_long %>%
  mutate(cluster_name = cluster_names[as.character(cluster)]) %>%
  mutate(cluster_name = factor(cluster_name, levels = cluster_names))

# Plot
ggplot(lime_long, aes(x = interaction(sampling_type, cluster_name), y = value)) +
  geom_boxplot(aes(color = sampling_type), alpha = 0.3, outlier.shape = NA) +
  geom_jitter(aes(color = cluster_name), width = 0.2, alpha = 0.7, size = 1.5) +
  facet_wrap(~ nutrient, scales = "free_y") +
  labs(
    x = "Sampling Type and Region",
    y = "Log-Transformed Nutrient Value (except pH)",
    title = "Nutrient Values by Sampling Type and Region",
    color = "Region"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```


```{r ANOVA assumptions check, warning=FALSE}
# Check ANOVA assumptions: normality and homogeneity of variance
#Ca
# For normality
plot(aov(ca ~ type, data = lime))  # Residuals vs Fitted + Q-Q plot
# For equal variances
car::leveneTest(ca ~ type, data = lime)
#Mg
plot(aov(mg ~ type, data = lime))  # Residuals vs Fitted + Q-Q plot
car::leveneTest(mg ~ type, data = lime)
#Zn
plot(aov(zn ~ type, data = lime))  # Residuals vs Fitted + Q-Q plot
car::leveneTest(zn ~ type, data = lime)
```


```{r sampling analysis, echo=FALSE, message=FALSE, warning=FALSE}
#sample analysis
#compare o-4, A, and B samples by location using a statistical test for each nutrient:

# phosphorus (P)
print("Phosphorus (P) Analysis")
anova_p <- aov(p ~ type, data = lime)
summary(anova_p)
# potassium (K)
print("Potassium (K) Analysis")
anova_k <- aov(k ~ type, data = lime)
summary(anova_k)
# calcium (Ca)
print("Calcium (Ca) Analysis")
anova_ca <- aov(ca ~ type, data = lime)
summary(anova_ca)
# magnesium (Mg)
print("Magnesium (Mg) Analysis")
anova_mg <- aov(mg ~ type, data = lime)
summary(anova_mg)
# zinc (Zn)
print("Zinc (Zn) Analysis")
anova_zn <- aov(zn ~ type, data = lime)
summary(anova_zn)
# pH
print("pH Analysis")
anova_ph <- aov(p_h_2 ~ type, data = lime)
summary(anova_ph)
# lbc
print("Lime Buffer Capacity (LBC) Analysis")
anova_lbc <- aov(lbc_1 ~ type, data = lime)
summary(anova_lbc)
# lbc_eq
print("Lime Buffer Capacity Equivalent (LBC Eq) Analysis")
anova_lbc_eq <- aov(lb_ceq ~ type, data = lime)
summary(anova_lbc_eq)
```
```{r}
# Post-hoc tests for significant ANOVA results
print("Ca Post-hoc Test")
TukeyHSD(aov(ca ~ type, data = lime))
print("Mg Post-hoc Test")
TukeyHSD(aov(mg ~ type, data = lime))
print("P Post-hoc Test")
TukeyHSD(aov(zn ~ type, data = lime))

```
## Assumptions check:

```{r assumptions check, echo=FALSE, message=FALSE, warning=FALSE}
#check assummptions for t-test:
# Normality check
print("Normality Check for Nutrients by Sampling Type")

shapiro.test(lime$ca[lime$sampling_type == "C"])

shapiro.test(lime$ca[lime$sampling_type == "SP"])

shapiro.test(lime$mg[lime$sampling_type == "C"])

shapiro.test(lime$mg[lime$sampling_type == "SP"])

shapiro.test(lime$p[lime$sampling_type == "C"])
shapiro.test(lime$p[lime$sampling_type == "SP"])
shapiro.test(lime$zn[lime$sampling_type == "C"])
shapiro.test(lime$zn[lime$sampling_type == "SP"])
shapiro.test(lime$p_h_2[lime$sampling_type == "C"])
shapiro.test(lime$p_h_2[lime$sampling_type == "SP"])
shapiro.test(lime$lbc_1[lime$sampling_type == "C"])
shapiro.test(lime$lbc_1[lime$sampling_type == "SP"])
shapiro.test(lime$lb_ceq[lime$sampling_type == "C"])
shapiro.test(lime$lb_ceq[lime$sampling_type == "SP"])
shapiro.test(lime$k[lime$sampling_type == "C"])
shapiro.test(lime$k[lime$sampling_type == "SP"])



# Variance check
car::leveneTest(ca ~ sampling_type, data = lime)
car::leveneTest(mg ~ sampling_type, data = lime)
car::leveneTest(p ~ sampling_type, data = lime)
car::leveneTest(zn ~ sampling_type, data = lime)
car::leveneTest(p_h_2 ~ sampling_type, data = lime)
car::leveneTest(lbc_1 ~ sampling_type, data = lime)
car::leveneTest(lb_ceq ~ sampling_type, data = lime)
car::leveneTest(k ~ sampling_type, data = lime)



```



All nutrients fail normality in both C and SP groups (all p-values < 0.05).

Therefore, I should not use t-tests.

Most nutrients pass homogeneity of variance, except for one (possibly p_h_2 or k), which had p = 0.02083.

A test that takes into account for no normality:

## Wilcoxon Rank Sum Tests (Mann–Whitney U):

```{r Wilcoxon analysis}
# nutrients <- c("ca", "mg", "zn", "p", "k", "p_h_2", "lbc_1", "lb_ceq")
# 
# wilcox_results <- lapply(nutrients, function(nutrient) {
#   formula <- as.formula(paste(nutrient, "~ sampling_type"))
#   test <- wilcox.test(formula, data = lime)
#   
#   data.frame(
#     nutrient = nutrient,
#     W = test$statistic,
#     p_value = test$p.value,
#     median_C = median(lime[[nutrient]][lime$sampling_type == "C"], na.rm = TRUE),
#     median_SP = median(lime[[nutrient]][lime$sampling_type == "SP"], na.rm = TRUE)
#   )
# })
# 
# wilcox_df <- do.call(rbind, wilcox_results)
# wilcox_df

```


```{r Wilcoxon test with nested locations}

# Step 1: Average A and B to create one SP value per location
lime_avg <- lime %>%
  mutate(sampling_type = ifelse(type %in% c("A", "B"), "SP", sampling_type)) %>%
  group_by(location, sampling_type) %>%
  summarise(across(c(ca, k, mg, mn, p, zn, p_h_2, lbc_1, lb_ceq), mean, na.rm = TRUE), .groups = "drop")

# Step 2: Reshape so we can do paired comparisons
lime_wide <- lime_avg %>%
  pivot_wider(names_from = sampling_type, values_from = c(ca, k, mg, mn, p, zn, p_h_2, lbc_1, lb_ceq))

# Step 3: Paired Wilcoxon test for each nutrient
nutrients <- c("ca", "k", "mg", "mn", "p", "zn", "p_h_2", "lbc_1", "lb_ceq")

results <- map_dfr(nutrients, function(nutrient) {
  C_col <- lime_wide[[paste0(nutrient, "_C")]]
  SP_col <- lime_wide[[paste0(nutrient, "_SP")]]

  # Paired test
  test <- wilcox.test(C_col, SP_col, paired = TRUE, exact = FALSE)

  tibble(
    nutrient = nutrient,
    p_value = test$p.value,
    median_C = median(C_col, na.rm = TRUE),
    median_SP = median(SP_col, na.rm = TRUE),
    W = test$statistic
  )
})
results

```


Most nutrients (except K and pH) show statistically significant differences between C and SP, with higher medians in C samples.

This supports the idea that single points (SP) tend to underrepresent nutrient values compared to composite (C), which might influence lime or fertilizer recommendations.

pH is not significantly different between the types... interesting and potentially important for lime recommendations since pH is often the direct driver.

### Linear Mixed Effects Model:

```{r}



# Linear mixed-effects model to account for nested structure (calcium)

#sum to zero contrasts
#log:
lime_log<- lime %>%
  mutate(across(c(ca, k, mg, mn, p, zn, p_h_2, lbc_1, lb_ceq), log1p))

options(contrasts = c("contr.sum", "contr.poly"))

lmer_results <- map_dfr(nutrients, function(nutrient) {
  formula <- as.formula(paste(nutrient, "~ sampling_type + (1 | location)"))
  model <- lmer(formula, data = lime_log)
  
  anova_result <- anova(model)
  
  tibble(
    nutrient = nutrient,
    p_value = anova_result$`Pr(>F)`[1],
    F_value = anova_result$`F value`[1],
    df = anova_result$Df[1],
    median_C = median(lime_avg[[paste0(nutrient, "_C")]], na.rm = TRUE),
    median_SP = median(lime_avg[[paste0(nutrient, "_SP")]], na.rm = TRUE)
  )
})

# Check normality of residuals for each nutrient

lmer_results %>%
  group_by(nutrient) %>%
  do({
    model <- lmer(as.formula(paste((.$nutrient), "~ sampling_type + (1 | location)")), data = lime_log)
    shapiro_test <- shapiro.test(residuals(model))
    tibble(nutrient = .$nutrient, p_value = shapiro_test$p.value)
  })

lmer_results
```


### Boxplots:

```{r boxplots all nutrients}

lime %>%
  filter(sampling_type %in% c("C", "SP")) %>%
  pivot_longer(cols = c(ca, k, mg, mn, p, zn, p_h_2, lbc_1, lb_ceq),
               names_to = "nutrient", values_to = "value") %>%
  ggplot(aes(x = sampling_type, y = value, fill = sampling_type)) +
  geom_boxplot(outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.8) +
  facet_wrap(~ nutrient, scales = "free_y") +
  scale_fill_manual(values = c("C" = "#1f77b4", "SP" = "#ff7f0e")) +
  labs(x = "Sampling Type", y = "Value", title = "Comparison of Nutrient Levels by Sampling Type") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

```


#Perry Data:

```{r Load perry data, echo=FALSE, message=FALSE, warning=FALSE}}
# Load Perry data

perry <- read_csv("../data_do_not_push/perry_soil_points_surface.csv") %>%
  clean_names() %>%
  select(label, type, x, y, lbc_1, lb_ceq, p_h_2, ca, k, mg, mn, p, zn, geometry)
```

```{r Perry assumptions}

# Define nutrient vectors
nutrients_log <- c("ca", "k", "mg", "mn", "p", "zn", "lbc_1", "lb_ceq")
nutrients <- c(nutrients_log, "p_h_2")

# Log-transform all nutrients except pH
perry_log <- perry %>%
  mutate(across(all_of(nutrients_log), log1p))

# Normality: Shapiro-Wilk test for each nutrient and group
normality_results <- map_dfr(nutrients, function(nutrient) {
  dat <- perry_log
  c_val <- dat[[nutrient]][dat$type == "CID"]
  sp_val <- dat[[nutrient]][dat$type == "SPID"]
  tibble(
    nutrient = nutrient,
    group = c("CID", "SPID"),
    p_value = c(shapiro.test(c_val)$p.value, shapiro.test(sp_val)$p.value)
  )
})

print("Shapiro-Wilk Normality Test Results:")
print(normality_results)

# Homogeneity of variance: Levene's test for each nutrient
variance_results <- map_dfr(nutrients, function(nutrient) {
  dat <- perry_log
  res <- car::leveneTest(as.formula(paste0(nutrient, " ~ type")), data = dat)
  tibble(
    nutrient = nutrient,
    p_value = res$`Pr(>F)`[1]
  )
})

print("Levene's Test for Homogeneity of Variance:")
print(variance_results)
```

```{r}
#wilcoxon:
#log:
  # perry_log <- perry %>%
  #   mutate(across(c(ca, k, mg, mn, p, zn, p_h_2, lbc_1, lb_ceq), log1p))
perry_wide <- perry %>%
  select(label, type, all_of(nutrients)) %>%
  pivot_wider(names_from = type, values_from = all_of(nutrients))  # gives e.g. ca_CID and ca_SPID

wilcox_results <- map_dfr(nutrients, function(nutrient) {
  # Create the two paired vectors
  x <- perry_wide[[paste0(nutrient, "_CID")]]
  y <- perry_wide[[paste0(nutrient, "_SPID")]]

  # Remove any NA pairs
  complete_idx <- complete.cases(x, y)
  x <- x[complete_idx]
  y <- y[complete_idx]

  # Wilcoxon signed-rank test (paired)
  wt <- wilcox.test(x, y, paired = TRUE, exact = FALSE)

  # Output a row
  tibble(
    nutrient = nutrient,
    p_value = wt$p.value,
    median_CID = median(x, na.rm = TRUE),
    median_SPID = median(y, na.rm = TRUE),
    W = wt$statistic
  )
})
wilcox_results
```

```{r LMER}
#do same but with lmer 
#log:
  # perry_log <- perry %>%
  #   mutate(across(c(ca, k, mg, mn, p, zn, p_h_2, lbc_1, lb_ceq), log1p))
lmer_perry_results <- map_dfr(nutrients, function(nutrient) {
  formula <- as.formula(paste(nutrient, "~ type + (1 | label)"))
  model <- lmer(formula, data = perry)
  
  anova_result <- anova(model)
  
  tibble(
    nutrient = nutrient,
    p_value = anova_result$`Pr(>F)`[1],
    F_value = anova_result$`F value`[1],
    df = anova_result$Df[1],
    median_CID = median(perry_wide[[paste0(nutrient, "_CID")]], na.rm = TRUE),
    median_SPID = median(perry_wide[[paste0(nutrient, "_SPID")]], na.rm = TRUE)
  )
})
lmer_perry_results
```

### Formatting:
```{r}
# Helper function for significance stars
sig_stars <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ ""
  )
}

# Liming Wilcoxon table
liming_wilcox_table <- results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, median_C, median_SP, W, p_value, significance) %>%
  arrange(factor(nutrient, levels = nutrients))

# Liming LMER table
liming_lmer_table <- lmer_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, F_value, p_value, significance) %>%
  arrange(factor(nutrient, levels = nutrients))

# Perry Wilcoxon table
perry_wilcox_table <- wilcox_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, median_CID, median_SPID, W, p_value, significance) %>%
  rename(median_C = median_CID, median_SP = median_SPID) %>%
  arrange(factor(nutrient, levels = nutrients))

# Perry LMER table
perry_lmer_table <- lmer_perry_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, median_CID, median_SPID, F_value, p_value, significance) %>%
  rename(median_C = median_CID, median_SP = median_SPID) %>%
  arrange(factor(nutrient, levels = nutrients))

# Print all tables using knitr::kable for neat output
library(knitr)
kable(liming_wilcox_table, caption = "Liming Study: Wilcoxon Signed-Rank Test")
kable(liming_lmer_table, caption = "Liming Study: Linear Mixed-Effects Model (log)")
kable(perry_wilcox_table, caption = "Perry Study: Wilcoxon Signed-Rank Test")
kable(perry_lmer_table, caption = "Perry Study: Linear Mixed-Effects Model")
```

### Log transform and test:

```{r}
# ---- Define nutrient names ----
nutrients_log <- c("ca", "k", "mg", "mn", "p", "zn", "lbc_1", "lb_ceq")
nutrients <- c(nutrients_log, "p_h_2")  # pH remains untransformed

# ---- Log-transform nutrients except pH ----
lime_log <- lime %>%
  mutate(across(all_of(nutrients_log), log1p))

perry_log <- perry %>%
  mutate(across(all_of(nutrients_log), log1p))

# ---- Significance stars helper ----
sig_stars <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ ""
  )
}

# ---- Shapiro-Wilk test function ----
shapiro_cluster_check <- function(df, nutrient_name, group_var) {
  df %>%
    filter(!is.na(.data[[nutrient_name]])) %>%
    group_by(.data[[group_var]]) %>%
    summarise(
      p_value = shapiro.test(.data[[nutrient_name]])$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      nutrient = nutrient_name,
      group = .data[[group_var]],
      sig = sig_stars(p_value)
    )
}

# ---- Run normality tests ----
log_normality_lime <- map_dfr(nutrients, ~ shapiro_cluster_check(lime_log, .x, "sampling_type"))
log_normality_perry <- map_dfr(nutrients, ~ shapiro_cluster_check(perry_log, .x, "type"))

# ---- Display results ----
library(knitr)
kable(log_normality_lime, caption = "Shapiro-Wilk Normality Test (Log-Transformed Nutrients + Raw pH) — Lime")
kable(log_normality_perry, caption = "Shapiro-Wilk Normality Test (Log-Transformed Nutrients + Raw pH) — Perry")


# ---- Liming: Wilcoxon (paired, log except pH) ----
lime_avg_log <- lime_log %>%
  mutate(sampling_type = ifelse(type %in% c("A", "B"), "SP", sampling_type)) %>%
  group_by(location, sampling_type) %>%
  summarise(across(all_of(nutrients), mean, na.rm = TRUE), .groups = "drop")

lime_wide_log <- lime_avg_log %>%
  pivot_wider(names_from = sampling_type, values_from = all_of(nutrients))

liming_wilcox_results <- map_dfr(nutrients, function(nutrient) {
  C_col <- lime_wide_log[[paste0(nutrient, "_C")]]
  SP_col <- lime_wide_log[[paste0(nutrient, "_SP")]]
  test <- wilcox.test(C_col, SP_col, paired = TRUE, exact = FALSE)
  tibble(
    nutrient = nutrient,
    p_value = test$p.value,
    median_C = median(C_col, na.rm = TRUE),
    median_SP = median(SP_col, na.rm = TRUE),
    W = test$statistic
  )
})

# ---- Liming: LMER (log except pH) ----
liming_lmer_results <- map_dfr(nutrients, function(nutrient) {
  formula <- as.formula(paste(nutrient, "~ sampling_type + (1 | location)"))
  model <- lmer(formula, data = lime_log)
  anova_result <- anova(model)
  tibble(
    nutrient = nutrient,
    p_value = anova_result$`Pr(>F)`[1],
    F_value = anova_result$`F value`[1]
  )
})

# ---- Perry: Wilcoxon (paired, log except pH) ----
perry_wide_log <- perry_log %>%
  select(label, type, all_of(nutrients)) %>%
  pivot_wider(names_from = type, values_from = all_of(nutrients))

perry_wilcox_results <- map_dfr(nutrients, function(nutrient) {
  x <- perry_wide_log[[paste0(nutrient, "_CID")]]
  y <- perry_wide_log[[paste0(nutrient, "_SPID")]]
  complete_idx <- complete.cases(x, y)
  x <- x[complete_idx]
  y <- y[complete_idx]
  wt <- wilcox.test(x, y, paired = TRUE, exact = FALSE)
  tibble(
    nutrient = nutrient,
    p_value = wt$p.value,
    median_CID = median(x, na.rm = TRUE),
    median_SPID = median(y, na.rm = TRUE),
    W = wt$statistic
  )
})

# ---- Perry: LMER (log except pH) ----
perry_lmer_results <- map_dfr(nutrients, function(nutrient) {
  formula <- as.formula(paste(nutrient, "~ type + (1 | label)"))
  model <- lmer(formula, data = perry_log)
  anova_result <- anova(model)
  tibble(
    nutrient = nutrient,
    p_value = anova_result$`Pr(>F)`[1],
    F_value = anova_result$`F value`[1]
  )
})

# ---- Format and print all tables ----
liming_wilcox_table <- liming_wilcox_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, median_C, median_SP, W, p_value, significance) %>%
  arrange(factor(nutrient, levels = nutrients))

liming_lmer_table <- liming_lmer_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, F_value, p_value, significance) %>%
  arrange(factor(nutrient, levels = nutrients))

perry_wilcox_table <- perry_wilcox_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, median_CID, median_SPID, W, p_value, significance) %>%
  rename(median_C = median_CID, median_SP = median_SPID) %>%
  arrange(factor(nutrient, levels = nutrients))

perry_lmer_table <- perry_lmer_results %>%
  mutate(significance = sig_stars(p_value)) %>%
  select(nutrient, F_value, p_value, significance) %>%
  arrange(factor(nutrient, levels = nutrients))

kable(liming_wilcox_table, caption = "Liming Study: Wilcoxon Signed-Rank Test (log-transformed except pH)")
kable(liming_lmer_table, caption = "Liming Study: Linear Mixed-Effects Model (log-transformed except pH)")
kable(perry_wilcox_table, caption = "Perry Study: Wilcoxon Signed-Rank Test (log-transformed except pH)")
kable(perry_lmer_table, caption = "Perry Study: Linear Mixed-Effects Model (log-transformed except pH)")
```


